The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-04 script/en/chapters/numbertheory.tex
reduced number of pdfTeX warnings (hyperlink related) from 615 to 15; moved index{...} out of headings
CT script, chap 4: Sage samples completed (en)

% --------------------------------------------------------------------------
\subsection{The Euler function\index{Euler!(phi) function}}
\label{L-Euler-Function}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}
} function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsection{Fixpoints\index{Fixpoint} modulo 26}

@@

The number $2^{1039}-1$ consists of 3 prime factors: The smallest one, 
$p7 = 5080711$ was already known.\footnote{%
This one can also be found using CrypTool\index{CrypTool}  via menu 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number} --- with the algorithms of Brent, Williams or
Lenstra, which are ``relativ'' good to separate small factors. \\
CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
}

@@

% ---------------------------------------------------------------------------
\hypertarget{AppArith1}{}
\subsection{Multiplication table modulo $m$}
\label{l:AppArith1}{}

@@

If $p$ is prime, then $\mathbf{Z} / p \mathbf{Z}$ has at least one
primitive root. Sometimes we want to compute all the primitive roots
of $\mathbf{Z} / p \mathbf{Z}$, not just any primitive root of
$\mathbf{Z} / p \mathbf{Z}$. The following function can do this.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_an_Integer(M):
....:     r"""
....:     Return all the primitive roots of M.
....:     """
....:     g = primitive_root(M)
....:     targetOrder = euler_phi(M)
....:     L = []
....:     # Stepping through all odd integers from 1 up to M, not including
....:     # M. So this loop only considers values of i where 1 <= i < M.
....:     for i in xrange(1, M, 2):
....:         testGen = Mod(g^i, M)
....:         if testGen.multiplicative_order() == targetOrder:
....:             L.append(testGen.lift())
....:     # removing duplicates
....:     return Set(L)

###aa
Alternativ-Code: In Def. abfangen, wenn die Fkt. primitive_root nicht berechenbar. Tut noch nicht.
sage: def enum_PrimitiveRoots_of_an_Integer(M):
	r"""
	Return all the primitive roots of M.
	"""
	try:
		g = primitive_root(M)
		print "Das ist %s" % M
	except:
		# raise ArithmeticError, "primitive_root() failed for %s." % M
		raise ArithmeticError, "primitive_root() failed."
	targetOrder = euler_phi(M)
	L = []
	# Stepping through all odd integers from 1 up to M, not including
	# M. So this loop only considers values of i where 1 <= i < M.
	for i in xrange(1, M, 2):
		testGen = Mod(g^i, M)
		if testGen.multiplicative_order() == targetOrder:
			L.append(testGen.lift())
	# removing duplicates
	return Set(L)

###zz

\end{Verbatim}
\caption{Function to calculate all primitive roots for a given number}
\end{sagecode}

@@

%########## Find number of primes and primitive roots in given range and time
\noindent With a little bit of programming, we can count how many primes are in a given range
and enumerate all their primitive roots.

With time command we can find out how long this takes.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_a_Range_of_Primes(start, end):
....:      r"""
....:      Compute all primitive roots of all primes between start and end,
....:      inclusive. This uses a primes iterator.
....:      """
....:      nPrimes = 0
....:      nPrimitiveRoots = 0
....:      for p in primes(start, end+1):
....:              nPrimes += 1
....:          L = enum_PrimitiveRoots_of_an_Integer(p)
....:          nPrimitiveRoots += len(L)
....:      print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nPrimes
....:
sage: time primroots_iter_tell_primenumber(1, 1500)
Enumerate all 62044 primitive roots of 239 primes.
CPU times: user 6.46 s, sys: 0.18 s, total: 6.64 s
Wall time: 7.48 s
\end{Verbatim}
\caption{Function to calculate all primitive roots for some given primes}
\end{sagecode}

%###### Find for all numbers if possible, switch between primes and numbers to select
\noindent And we can count how many primes for which to enumerate all
their primitive roots.

%a##########################
\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_a_Range_of_Integers(start, end, bPrimesOnly=True):
	r"""
	Compute all primitive roots of all primes between start and end,
	inclusive. This actually performs primality tests, so it allows us
	to count the number of primes from start to end, inclusive.
	"""
	nCheckedNumb = 0
	nPrimitiveRoots = 0
	for n in xrange(start, end+1):
		if bPrimesOnly:
			if is_prime(n):
				nCheckedNumb += 1
				L = enum_PrimitiveRoots_of_an_Integer(n)
				nPrimitiveRoots += len(L)
		else:
			nCheckedNumb += 1
			L = enum_PrimitiveRoots_of_an_Integer(n)
			nPrimitiveRoots += len(L)
	if bPrimesOnly:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nCheckedNumb
	else:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s numbers." % nCheckedNumb

sage: %time primroots_counter(1, 500)

sage: %time primroots_counter(1, 500, False)

Problem z.B. beim Wert 8.  Abfangen per try - except, wenn keine primitive root mod m ex. !

%b##########################

sage: def primroots_primetest(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This actually performs primality tests, so it allows us
....:     to count the number of primes from start to end, inclusive.
....:     """
....:     nPrimes = 0
....:     for n in xrange(start, end+1):
....:         if is_prime(n):
....:             nPrimes += 1
....:             enum_primroots(n)
....:     print "Enumerate all primitive roots of %s primes." % nPrimes
....:     
sage: %time primroots_primetest(1, 1500)
Enumerate all primitive roots of 239 primes.
CPU times: user 2.35 s, sys: 0.03 s, total: 2.38 s
Wall time: 2.39 s
\end{Verbatim}
\caption{Another function to calculate all primitive roots for some given primes}
\end{sagecode}

\newpage
Using our custom-defined function \verb!enum_primroots!, we can find
all primitive roots of a prime integer $p$. From this list of primitive
roots, we can determine the smallest and largest primitive root for
$\mathbf{Z} / p \mathbf{Z}$, as well as count the number of primitive
roots of $\mathbf{Z} / p \mathbf{Z}$.
Figure~\ref{fig:primitive_roots_all} graphs the number of primitive
roots for each prime between 1 and 100,000. The $x$-axis represents
primes between 1 and 100,000, while the $y$-axis counts the number of
primitive roots for each prime within that interval.

@@

Figure~\ref{fig:primitive_roots_smallest} graphs the smallest
primitive roots of all primes between 1 and 100,000. The $x$-axis
represents primes between 1 and 100,000. The $y$-axis represents the
smallest primitive root of each prime within that
interval. Figure~\ref{fig:primitive_roots_largest} shows a
corresponding graph for the largest primitive root of each prime
within the above interval.

@@

% To enforce that the next chapter starts at the beginning of a new page after
% the freely placed graphics, it is necessary to have TWO newpages and in between
% something to be printed (here invisible blanks).
\newpage
$ $

@@

  \item {Mr.\ Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed in his leisure time the 
	 functions and the complex dialog box of the RSA cryptosystem, 
	 which enables you to execute the RSA samples of this article.}

  \item {J\"org Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}

  \item {Dr. Georg Illies for pointing me to Pari-GP \index{Pari-GP}.}

>>>>

% --------------------------------------------------------------------------
\subsection{The Euler function}\index{Euler!(phi) function}
\label{L-Euler-Function}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}
} function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsection{Fixpoints modulo 26}\index{Fixpoint}

@@

The number $2^{1039}-1$ consists of 3 prime factors: The smallest one, 
$p7 = 5080711$ was already known.\footnote{%
This one can also be found using CrypTool\index{CrypTool}  via menu 
{\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem \textbackslash{} 
Factorization of a Number} --- with the algorithms of Brent, Williams or
Lenstra, which are ``relatively'' good to separate small factors. \\
CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
}

@@

% ---------------------------------------------------------------------------
\hypertarget{AppArith1}{}
\subsection{Multiplication table modulo m}     % $ removed at $m$
\label{l:AppArith1}{}

@@

\noindent If $p$ is prime, then $\mathbf{Z} / p \mathbf{Z}$ has at least one
primitive root.

\newpage
Sometimes we want to compute all the primitive roots
of $\mathbf{Z} / p \mathbf{Z}$, not just any primitive root of
$\mathbf{Z} / p \mathbf{Z}$.
The following function can do this%
\footnote{This code was developed in a Sage script file and
executed non-interactively. That is why you don't see "sage:" and "....:"
at the beginning of the lines like in the Sage samples before.}.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def enum_PrimitiveRoots_of_an_Integer(M):
    r"""
    Return all the primitive roots of the integer M (if possible).
    """
    try:
        g = primitive_root(M)
    except:
        return None
    targetOrder = euler_phi(M)
    L=[]
    # Stepping through all odd integers from 1 up to M, not including
    # M. So this loop only considers values of i where 1 <= i < M.
    for i in xrange(1,M,2):
            testGen = Mod(g^i,M)
            if testGen.multiplicative_order() == targetOrder:
                L.append(testGen.lift())
    # removing duplicates
    return Set(L)

# AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)
print "AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)"
M=10; print "1-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=8; print "2-----------Testcase: M = %s" % M
# M=8 hat keine primitive root mod m. Checke, ob per try - except abgefangen.
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
M=17; print "3-----------Testcase: M = %s" % M
LL = enum_PrimitiveRoots_of_an_Integer(M)
if LL==None:
    print M
else:
    print LL
# AA_End -- Testcases

OUTPUT:
AA_Start -- Testcases for enum_PrimitiveRoots_of_an_Integer(M)
1-----------Testcase: M = 10
{3, 7}
2-----------Testcase: M = 8
8
3-----------Testcase: M = 17
{3, 5, 6, 7, 10, 11, 12, 14}
\end{Verbatim}
\caption{Function ``enum\_PrimitiveRoots\_of\_an\_Integer'' to calculate all primitive roots for a given number}
\end{sagecode}

@@

\newpage
\noindent With a little bit of programming, we can count how many primitive roots
are in a given range of integers. We can check this for all numbers or only for the
primes within this range

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True):
	r"""
	Compute all primitive roots of all numbers between start and end,
	inclusive, and count them.
	If the flag bPrimesOnly is True, it performs primality tests, so it
	allows us to count the number of primes from start to end, inclusive.
        If the flag bPrimesOnly is false, it additionally counts these even
	numbers which have no primitive root.
	"""
	nCheckedNumb = 0
	nCheckedNumb_WithoutPrimitivRoots = 0
	nPrimitiveRoots = 0
	for n in xrange(start, end+1):
		if bPrimesOnly:
			if is_prime(n):
				nCheckedNumb += 1
				L = enum_PrimitiveRoots_of_an_Integer(n)
				nPrimitiveRoots += len(L)
		else:
			nCheckedNumb += 1
			L = enum_PrimitiveRoots_of_an_Integer(n)
			if L==None:
				nCheckedNumb_WithoutPrimitivRoots += 1
			else:
				nPrimitiveRoots += len(L)

	if bPrimesOnly:
		print "Found all %s" % nPrimitiveRoots + \
		      " primitive roots of %s primes." % nCheckedNumb
	else:
		if nCheckedNumb_WithoutPrimitivRoots == 0:
			print "Found all %s " % nPrimitiveRoots + \
			      "primitive roots of %s numbers." % nCheckedNumb
		else:
			print "Found all %s " % nPrimitiveRoots + \
			      "primitive roots of %s numbers." % \
			          (nCheckedNumb - nCheckedNumb_WithoutPrimitivRoots)
			print "(Total of numbers checked: %s, " % nCheckedNumb + \
			      "Amount of numbers without primitive roots: %s)" % \
			          nCheckedNumb_WithoutPrimitivRoots
\end{Verbatim}
\caption{Function ``count\_PrimitiveRoots\_of\_an\_IntegerRange'' to calculate all primitive roots for a given range of integers}
\end{sagecode}

\newpage
\noindent Using the Sage command \verb!time!, we can also find out how long it takes on
our computer.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
# BB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)
print "\n\nBB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)"

print "\n1-----------Testcase: (1, 500)"
time count_PrimitiveRoots_of_an_IntegerRange(1, 500)

print "\n2-----------Testcase: (5, 6, False)"
time count_PrimitiveRoots_of_an_IntegerRange(5, 6, False)

print "\n3-----------Testcase: (1, 500, False)"
time count_PrimitiveRoots_of_an_IntegerRange(1, 500, False)
# BB_End -- Testcases

OUTPUT:
BB_Start -- Testcases for count_PrimitiveRoots_of_an_IntegerRange(start, end, bPrimesOnly=True)

1-----------Testcase: (1, 500)
Found all 8070 primitive roots of 95 primes.
Time: CPU 0.94 s, Wall: 0.97 s

2-----------Testcase: (5, 6, False)
Found all 3 primitive roots of 2 numbers.
Time: CPU 0.00 s, Wall: 0.00 s

3-----------Testcase: (1, 500, False)
Found all 11010 primitive roots of 170 numbers.
(Total of numbers checked: 500, Amount of numbers without primitive roots: 330)
Time: CPU 1.52 s, Wall: 1.59 s
\end{Verbatim}
\caption{Testcases and output of function ``count\_PrimitiveRoots\_of\_an\_IntegerRange''}
\end{sagecode}

\newpage
\newpage
\noindent Using our custom-defined function \verb!enum_PrimitiveRoots_of_an_Integer!,
we can find all primitive roots of one prime integer $p$.

The following function counts how many primes are in a given range
and enumerate all their primitive roots.

From this list of primitive roots, we can determine the smallest and largest
primitive root for $\mathbf{Z} / p \mathbf{Z}$, as well as count the number
of primitive roots of $\mathbf{Z} / p \mathbf{Z}$.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def count_PrimitiveRoots_of_a_PrimesRange(start, end):
      r"""
      Compute all primitive roots of all primes between start and end,
      inclusive. This uses a primes iterator.
      """
      nPrimes = 0
      nPrimitiveRoots = 0
      for p in primes(start, end+1):
          L = enum_PrimitiveRoots_of_an_Integer(p)
	  print p, len(L)
          nPrimes += 1
          nPrimitiveRoots += len(L)
      print "Found all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nPrimes

# CC_Start -- Testcases for count_PrimitiveRoots_of_a_PrimesRange(start, end)
print "\n\nBB_Start -- Testcases for count_PrimitiveRoots_of_a_PrimesRange(start, end)"
print "-----------Testcase: (1, 1500)"
time count_PrimitiveRoots_of_a_PrimesRange(1, 1500)
# CC_End -- Testcases

OUTPUT:
CC_Start -- Testcases for count_PrimitiveRoots_of_a_PrimesRange(start, end)
-----------Testcase: (1, 1500)
2 1
3 1
5 2
7 2
11 4
13 4
17 8
19 6
23 10
29 12
31 8
37 12
...
1483 432
1487 742
1489 480
1493 744
1499 636
Found all 62044 primitive roots of 239 primes.
Time: CPU 7.55 s, Wall: 7.85 s
\end{Verbatim}
\caption{Function to calculate all primitive roots for a given range of primes}
\end{sagecode}

\newpage
\newpage
\noindent A slightly modified version of our custom-defined function
\verb!count_PrimitiveRoots_of_a_PrimesRange!, is used to generate
a database of all primitive roots of all primes between 1 and 100,000.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
start = 1
end = 10^5
fileName = "/scratch/mvngu/primroots.dat"
file = open(fileName, "w")
for p in primes(start, end+1):
    L = enum_PrimitiveRoots_of_an_Integer(p)
    print p, len(L)
    # Output to a file. The format is:
    # (1) the prime number p under consideration
    # (2) the number of primitive roots of Z/pZ
    # (3) all the primitive roots of Z/pZ
    file.write(str(p) + " " + str(len(L)) + " " + str(L) + "\n")
    file.flush()
file.close()
\end{Verbatim}
\caption{Code to generate the database with all primitive roots for all primes between 1 and 100,000}
\end{sagecode}

It took about one day on the machine sage.math to generate the
file ``primroots.dat'' (done in July 2009 by Minh Van Nguyen).

This code and the function \verb!enum_PrimitiveRoots_of_an_Integer!
was put in a Sage script file and executed non-interactively.

The file ``primroots.dat'' is a database of all primitive roots of all primes
between 1 and 100,000 inclusive. It is a very large file (about 1 GB
uncompressed, and 285 MB compressed with bzip2). You can find the file
at
   \href{http://sage.math.washington.edu/home/mvngu/doc/primitive-roots/primroots.dat.bz2}
   {\url{http://sage.math.washington.edu/home/mvngu/doc/primitive-roots/primroots.dat.bz2}}.

\newpage
\newpage
\noindent This database file was used then to create three graphics using the following code.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # open a database file on primitive roots from 1 to 100,000
sage: file = open("/scratch/mvngu/primroots.dat", "r")
sage: plist = []    # list of all primes from 1 to 100,000
sage: nlist = []    # number of primitive roots modulo prime p
sage: minlist = []  # smallest primitive root modulo prime p
sage: maxlist = []  # largest primitive root modulo prime p
sage: for line in file:
....:     # get a line from the database file and tokenize it for processing
....:     line = line.strip().split(" ", 2)
....:     # extract the prime number p in question
....:     plist.append(Integer(line[0]))
....:     # extract the number of primitive roots modulo p
....:     nlist.append(Integer(line[1]))
....:     # extract the list of all primitive roots modulo p
....:     line = line[-1]
....:     line = line.replace("{", "")
....:     line = line.replace("}", "")
....:     line = line.split(", ")
....:     # sort the list in non-decreasing order
....:     line = [Integer(s) for s in line]
....:     line.sort()
....:     # get the smallest primitive root modulo p
....:     minlist.append(line[0])
....:     # get the largest primitive root modulo p
....:     maxlist.append(line[-1])
....:
sage: file.close()  # close the database file
sage: # plot of number of primitive roots modulo p
sage: nplot = point2d(zip(plist, nlist), pointsize=1)
sage: nplot.axes_labels(["x", "y"])
sage: nplot
sage: # plot of smallest primitive root modulo prime p
sage: minplot = point2d(zip(plist, minlist), pointsize=1)
sage: minplot.axes_labels(["x", "y"])
sage: minplot
sage: # plot of largest primitive root modulo prime p
sage: maxplot = point2d(zip(plist, maxlist), pointsize=1)
sage: maxplot.axes_labels(["x", "y"])
sage: maxplot
\end{Verbatim}
\caption{Code to generate the the graphics about the primitive roots}
\end{sagecode}

\newpage
Figure~\ref{fig:primitive_roots_all} graphs the number of primitive
roots for each prime between 1 and 100,000. The $x$-axis represents
primes between 1 and 100,000, while the $y$-axis counts the number of
primitive roots for each prime within that interval.

@@

\vskip +50 pt
Figure~\ref{fig:primitive_roots_smallest} graphs the smallest
primitive roots of all primes between 1 and 100,000. The $x$-axis
represents primes between 1 and 100,000. The $y$-axis represents the
smallest primitive root of each prime within that
interval.

\vskip +25 pt
Figure~\ref{fig:primitive_roots_largest} shows a
corresponding graph for the largest primitive root of each prime
within the above interval.

@@

% To enforce that the next chapter starts at the beginning of a new page after
% the freely placed graphics, it is necessary to have TWO newpages and in between
% something to be printed (here invisible blanks).
% \newpage
% $ $

@@

  \item {Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading of this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed in his leisure time the 
	 functions and the complex dialog box of the RSA cryptosystem within CrypTool, 
	 which enables you to execute the RSA samples of this article.}

  \item {J\"org Cornelius Schneider for his enthusiastic TeX support and for the
         many cases where he helped when facing programming or design problems.}

  \item {Dr.\ Georg Illies for pointing me to Pari-GP\index{Pari-GP}.}

  \item {Lars Fischer for his help with fast Pari-GP\index{Pari-GP} code for primitive roots.}

  \item {Minh Van Nguyen from Australia for his always fast and exhaustive
         help with the Sage\index{Sage} code samples in this chapter.}
